`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2016-04-28 response-to-reviewers.tex

Provide a better description of the stepwise analysis of fossil ages. My impression (but I am not totally sure) is that in one analysis, fossil ages are unknown parameters and have uniform prior distributions set according to the stratigraphic bounds. Then, in stepwise analysis, each fossil age is in turn represented not by the stratigraphic uniform distribution, but by a prior density derived from the fossilised birth-death process. Give some details please. I was trying to work out the general form of the posterior distribution, I was guessing something like:

$$p(tip\_ages) p(node\_ages) L(D|ages,tree)$$

where $p(tip\_ages)$ is the prior density on fossil ages using uniform distributions, $p(node\_ages)$ is the prior on internal node ages given the BD process, and $L(D|ages,tree)$ is the likelihood of the morphological data D, given the ages and the topology calculated using Lewis (2001) formula. However, it seems from the text in p.5 that another likelihood term may be involved. For the stepwise deletion procedure, I am guessing the priors are something like $p(tip\_ages\hat{}) p(node\_ages, tip\_age*)$, where now $tip\_ages\hat{}$ includes all the fossil ages except the one under analysis, which has been moved into the other BD-based prior term $(tip\_age*)$. Please provide general forms for the posteriors and make the connection with equation (1).

\response{We have now provided a general form of the posterior we are sampling under the section on ``Phylogenetic analyses'' and clarified details of the stepwise analysis of the fossils.}

@@

Give some more info on how the prior on the root age is set. Is Tmax the same as the root age? If it is, please state explicitly.

\response{$T$ is not the age of the root, it is the upper limit on the age of the root.}

@@

Please give the prior on the morphological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).

\response{We have added the prior information for the morphological substitution rate and log standard deviation of the relaxed clock.}

@@

p.8 paragraph starting on line 43. Do you mean that with the particular implementation of marginal likelihood calculation in BEAST you cannot calculate the BF? Or is this a general problem with any implementation of stepping-stones and this model? I am also confused, do sampling times refer to the ages of fossil tips? Aren't these unknown parameters being estimated in the MCMC? Then they would appear as terms in the prior and the likelihood. (?)

\response{The ages of the fossil tips are indeed unknown parameters estimated by the MCMC. Their prior distribution emerges from the marginal distribution of the fossilized birth death prior of tip-timed trees, which can be sampled using MCMC, but which we don't have an analytical solution for.}

### 2016-04-28 fossilDating-revision.tex
answering reviewer 1 detailed points

%TSrev
Given the morphological data $D$ and a stratographic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratographic age range. The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the morpholocical evolution model are summarized in $\theta$. More formally, we sample from,
$$P[\mathcal{T},\eta, \theta | D, a] = P[D|\mathcal{T},\theta] P[\mathcal{T},a|\eta] P[\eta] P[\theta]/P[D,a],$$
where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratographic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
When we replaced the focal  palaeontological age constraints of fossil $i$  by 
the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$.

@@

We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}. \Mstrict{} assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites. Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.

>>>>

%TSrev
Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range. The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the model for morphological character evolution are summarized in $\theta$. More formally, we sample from,
$$P[\mathcal{T},\eta, \theta | D, a] = P[D|\mathcal{T},\theta] P[\mathcal{T},a|\eta] P[\eta] P[\theta]/P[D,a],$$
where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratigraphic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
When we replaced the focal  palaeontological age constraints of fossil $i$  by 
the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$.

@@

We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}. \Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites. Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa. Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates. We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
